---
title: "Prediction of Cardiovascular and Healthcare Utilization Outcomes in Type 2 Diabetes"
author: "Your Name"
format:
  pdf:
    toc: true
    number-sections: true
    code-fold: false
execute:
  echo: true
  warning: false
  message: false
---

Due to substantial missing data in all the laboratory variables, I performed a complete-case analysis. I first conducted univariable analyses to explore associations between each predictor and the outcomes. I then fitted multivariable models, including a primary model with variables with less than 90% missing data and a secondary exploratory model including all laboratory variables (except from homocystein), recognizing the reduced sample size.

## 1. Load packages

```{r}
library(tidyverse)
library(naniar)
library(gtsummary)
library(broom)
library(pROC)
library(ResourceSelection)
```

## 2. Load data

```{r}
data1 <- read.csv("data1.csv", stringsAsFactors = FALSE)
summary_data1 <- read.csv("summary_data1.csv", stringsAsFactors = FALSE)

dim(data1)
```

## 3. Clean data

```{r}
data_clean <- data1 %>%
  distinct() %>%
  mutate(across(where(is.character), ~ na_if(.x, "censored")))
```

## 4. Define cohort

Restriction of the dataset to the study population of interest:

adults aged 18 years or older
patients with a diagnosis of Type 2 diabetes mellitus
diagnosis occurring during an emergency or inpatient encounter

This defines the final analytic cohort.

```{r}
cohort <- data_clean %>%
  filter(
    r_age_at_event >= 18,
    !is.na(r_condition),
    str_detect(r_condition, regex("Type 2 diabetes mellitus", ignore_case = TRUE)),
    r_encounter_types %in% c("emergency", "inpatient", "inpatient|emergency")
  )

nrow(cohort)
```

## 5. Create variables

baseline predictors: age, sex, and laboratory values measured within 90 days after the event
Outcome 1: stroke, heart failure, or chronic kidney failure within 10 years
Outcome 2: emergency department visit within 1 year

I also create factor versions of the outcomes for the descriptive tables.

```{r}
data <- cohort %>%
  mutate(
    age = r_age_at_event,
    sex = factor(gender, levels = c("Female", "Male")),

    hba1c = l_90dPost_hba1c_result_numeric_calculated,
    cholesterol = l_90dPost_cholesterol_total_result_numeric_calculated,
    glucose = l_90dPost_glucose_result_numeric_calculated,
    homocysteine = l_90dPost_homocystine_result_numeric_calculated,
    triglycerides = l_90dPost_triglycerides_result_numeric_calculated,
    creatinine = l_90dPost_creatinine_result_numeric_calculated,

    stroke_10y = if_else(
      !is.na(c_20yPost_avc_like_date_time_of_condition_assigned_days_from_reference) &
        c_20yPost_avc_like_date_time_of_condition_assigned_days_from_reference > 0 &
        c_20yPost_avc_like_date_time_of_condition_assigned_days_from_reference <= 3652,
      1, 0
    ),

    hf_10y = if_else(
      !is.na(c_20yPost_congestive_heart_failure_date_time_of_condition_assigned_days_from_reference) &
        c_20yPost_congestive_heart_failure_date_time_of_condition_assigned_days_from_reference > 0 &
        c_20yPost_congestive_heart_failure_date_time_of_condition_assigned_days_from_reference <= 3652,
      1, 0
    ),

    ckd_10y = if_else(
      !is.na(c_20yPost_chonic_kidney_disease_date_time_of_condition_assigned_days_from_reference) &
        c_20yPost_chonic_kidney_disease_date_time_of_condition_assigned_days_from_reference > 0 &
        c_20yPost_chonic_kidney_disease_date_time_of_condition_assigned_days_from_reference <= 3652,
      1, 0
    ),

    outcome1 = if_else(
      stroke_10y == 1 | hf_10y == 1 | ckd_10y == 1,
      1, 0
    ),

    outcome2 = if_else(
      !is.na(e_30dPost_admission_start_date_days_from_reference) &
        e_30dPost_admission_start_date_days_from_reference > 0 &
        e_30dPost_admission_start_date_days_from_reference <= 365,
      1, 0
    ),

    outcome1_f = factor(outcome1, levels = c(0, 1), labels = c("No", "Yes")),
    outcome2_f = factor(outcome2, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

## 6. Descriptive statistics

This gives a quick overview of the final cohort:

total sample size
mean age
proportion of males
frequency of Outcome 1 and Outcome 2
```{r}
data %>%
  summarise(
    N = n(),
    age_mean = mean(age, na.rm = TRUE),
    male_pct = mean(sex == "Male", na.rm = TRUE) * 100,
    outcome1_pct = mean(outcome1, na.rm = TRUE) * 100,
    outcome2_pct = mean(outcome2, na.rm = TRUE) * 100
  )
```

## 7. Missing data

 Before modeling, I examine missingness in the predictors.
This is important because the regression models below use complete-case analysis, meaning that only patients with complete data for the variables in each model are included.

```{r}
data %>%
  summarise(across(c(age, sex, hba1c, cholesterol, glucose,
                     homocysteine, triglycerides, creatinine),
                   ~ mean(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_prop")
```

## 8. Table 1 for Outcome 1

This table compares baseline characteristics between patients with and without Outcome 1.

```{r}
data %>%
  select(outcome1_f, age, sex, hba1c, cholesterol,
         glucose, triglycerides, creatinine) %>%
  tbl_summary(
    by = outcome1_f,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})"
    )
  ) %>%
  add_p()
```

## 9. Table 1 for Outcome 2

This table compares baseline characteristics between patients with and without Outcome 2.

```{r}
data %>%
  select(outcome2_f, age, sex, hba1c, cholesterol,
         glucose, triglycerides, creatinine) %>%
  tbl_summary(
    by = outcome2_f,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})"
    )
  ) %>%
  add_p()
```

## 10. Univariable logistic regression function

I now define a function to run univariable logistic regression.
This means that each predictor is tested one at a time against the outcome.

The function:

keeps only the outcome and one predictor
removes rows with missing values
fits a logistic regression model
returns odds ratios, confidence intervals, and p-values

```{r}
vars <- c("age", "sex", "hba1c", "cholesterol",
          "glucose", "homocysteine",
          "triglycerides", "creatinine")

run_uni <- function(outcome_var, predictor_var, data) {
  d <- data %>%
    select(all_of(outcome_var), all_of(predictor_var)) %>%
    drop_na()

  if (nrow(d) < 50) return(NULL)

  fit <- glm(
    as.formula(paste(outcome_var, "~", predictor_var)),
    data = d,
    family = binomial()
  )

  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      outcome = outcome_var,
      variable = predictor_var,
      n = nrow(d)
    ) %>%
    select(outcome, variable, n, estimate, conf.low, conf.high, p.value)
}
```

## 11. Univariable logistic regression for Outcome 1

Here I run separate univariable models for each predictor using Outcome 1.

```{r}
uni_outcome1 <- map_dfr(vars, ~ run_uni("outcome1", .x, data))

uni_outcome1
```


## 12. Multivariable model for Outcome 1: primary model

This is the main multivariable model for Outcome 1.
It includes:

age
sex
glucose
creatinine

These predictors are included because they are clinically relevant and have better data availability than some of the other laboratory variables.

```{r}
main_data_outcome1 <- data %>%
  select(outcome1, age, sex, glucose, creatinine) %>%
  drop_na()

model_main_outcome1 <- glm(
  outcome1 ~ age + sex + glucose + creatinine,
  data = main_data_outcome1,
  family = binomial()
)

broom::tidy(model_main_outcome1, exponentiate = TRUE, conf.int = TRUE)
```

## 13. Multivariable model for Outcome 1: full lab model
This is a secondary, more exploratory model for Outcome 1 that additionally includes:

HbA1c
cholesterol
triglycerides

Because these variables have more missing data, this model is fit on a smaller complete-case subset.

```{r}
full_data_outcome1 <- data %>%
  select(outcome1, age, sex, hba1c, cholesterol,
         glucose, triglycerides, creatinine) %>%
  drop_na()

model_full_outcome1 <- glm(
  outcome1 ~ age + sex + hba1c + cholesterol +
    glucose + triglycerides + creatinine,
  data = full_data_outcome1,
  family = binomial()
)

broom::tidy(model_full_outcome1, exponentiate = TRUE, conf.int = TRUE)
```

## 14. Univariable logistic regression for Outcome 2

Here I repeat the univariable analyses, but now using Outcome 2.
```{r}
uni_outcome2 <- map_dfr(vars, ~ run_uni("outcome2", .x, data))

uni_outcome2
```


## 15. Multivariable model for Outcome 2: primary model

This is the main multivariable model for Outcome 2, again using age, sex, glucose, and creatinine.

```{r}
main_data_outcome2 <- data %>%
  select(outcome2, age, sex, glucose, creatinine) %>%
  drop_na()

model_main_outcome2 <- glm(
  outcome2 ~ age + sex + glucose + creatinine,
  data = main_data_outcome2,
  family = binomial()
)

broom::tidy(model_main_outcome2, exponentiate = TRUE, conf.int = TRUE)
```

## 16. Multivariable model for Outcome 2: full lab model

This is the more exploratory full laboratory model for Outcome 2.

```{r}
full_data_outcome2 <- data %>%
  select(outcome2, age, sex, hba1c, cholesterol,
         glucose, triglycerides, creatinine) %>%
  drop_na()

model_full_outcome2 <- glm(
  outcome2 ~ age + sex + hba1c + cholesterol +
    glucose + triglycerides + creatinine,
  data = full_data_outcome2,
  family = binomial()
)

broom::tidy(model_full_outcome2, exponentiate = TRUE, conf.int = TRUE)
```



Below I evaluate how well the models discriminate between patients with and without Outcome 1 and 2 using:

ROC curves
Area under the curve (AUC)

## 17. Model performance for Outcome 1


```{r}
p_main_outcome1 <- predict(model_main_outcome1, type = "response")
roc_main_outcome1 <- roc(main_data_outcome1$outcome1, p_main_outcome1)
auc(roc_main_outcome1)
plot(roc_main_outcome1)
```

```{r}
p_full_outcome1 <- predict(model_full_outcome1, type = "response")
roc_full_outcome1 <- roc(full_data_outcome1$outcome1, p_full_outcome1)
auc(roc_full_outcome1)
plot(roc_full_outcome1)
```

## 18. Model performance for Outcome 2

```{r}
p_main_outcome2 <- predict(model_main_outcome2, type = "response")
roc_main_outcome2 <- roc(main_data_outcome2$outcome2, p_main_outcome2)
auc(roc_main_outcome2)
plot(roc_main_outcome2)
```

```{r}
p_full_outcome2 <- predict(model_full_outcome2, type = "response")
roc_full_outcome2 <- roc(full_data_outcome2$outcome2, p_full_outcome2)
auc(roc_full_outcome2)
plot(roc_full_outcome2)
```

## 19. Calibration

Finally, I assess calibration of the primary models using the Hosmer-Lemeshow test.
This checks how well predicted probabilities agree with observed outcomes.

```{r}
hoslem.test(main_data_outcome1$outcome1, fitted(model_main_outcome1), g = 10)
```

```{r}
hoslem.test(main_data_outcome2$outcome2, fitted(model_main_outcome2), g = 10)
```

## 20. Optional: save results

```{r}
write.csv(uni_outcome1, "univariable_outcome1.csv", row.names = FALSE)
write.csv(uni_outcome2, "univariable_outcome2.csv", row.names = FALSE)
```

```{r}
write.csv(
  broom::tidy(model_main_outcome1, exponentiate = TRUE, conf.int = TRUE),
  "multivariable_main_outcome1.csv",
  row.names = FALSE
)

write.csv(
  broom::tidy(model_full_outcome1, exponentiate = TRUE, conf.int = TRUE),
  "multivariable_full_outcome1.csv",
  row.names = FALSE
)

write.csv(
  broom::tidy(model_main_outcome2, exponentiate = TRUE, conf.int = TRUE),
  "multivariable_main_outcome2.csv",
  row.names = FALSE
)

write.csv(
  broom::tidy(model_full_outcome2, exponentiate = TRUE, conf.int = TRUE),
  "multivariable_full_outcome2.csv",
  row.names = FALSE
)
```


